HO unification from object language to meta language

### 2024-03-31 paper/main.tex
printer generator for unif problems in tex

\newcommand{\printMinipage}[2]{
  \listlength{#2}
  \foreach \x [count=\i] in #2
    {
      \begin{minipage}{3cm}
        $#1_\i : \x$ %
      \end{minipage}
      \ifthenelse{\equal{\i}{\thellength}}{}{
        \ifthenelse{\isodd{\i}}{|}{\\}
      }
    }
}

\newcommand{\printEntry}[1]{
  \listlength{#1}%
  \ifthenelse{\equal{0}{\thellength}}{}{
    \item[Entry pb:] 
    \foreach \x [count=\i] in #1
      {%
        \begin{minipage}{8cm}%
          $\mathbb{P}_\i : \x$ %
        \end{minipage}%
        \ifthenelse{\equal{\i}{\thellength}}{}{\\}%
      }%
  }%
}
\newcommand{\printItemComp}[3]{
  \listlength{#3}%
  \ifthenelse{\equal{0}{\thellength}}{}{
    \item[#1] \printMinipage{#2}{#3}
  }%
}

\newcommand{\printAll}[4]{
  \noindent%
  \begin{description}[leftmargin=!,labelwidth=\widthof{\bfseries Mapping:}]
    \printEntry{#1}%
    \printItemComp{Compil:}{\mathbb{T}}{#2}%
    \printItemComp{Mapping:}{m}{#3}%
    \printItemComp{Links:}{l}{#4}
  \end{description}
}

@@

\section{Handling of \maybeeta}\label{sec:eta}
\marginpar{TODO: fix syntax}
$\eta$-reduction is an equivalence relation where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expansion of $t$. 
\printAll
  {{\elpiIn{flam x\ fapp [fuva X, x]} \Uo \elpiIn{fcon"f"}}}
  {{\lambda x. X'_x \Ue f}}
  {{\mapping{\elpiIn{X}}{X'}{1}}}
  {{}}
Following the implementation of \elpiIn{comp} given in
\cref{sec:compilation}, the unification problem $\mathbb{P}_1$ given 
above, is compiled into the unification problem $\mathbb{T}_1$. 
We can note that all the terms appearing in $\mathbb{P}_1$ are
unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathbb{T}_1$:
its left and right hand sides have both different rigid head:
$\lambda x.A'_x$ corresponds to the \Ho term \elpiIn{lam x\ app[con "f", x]} 
and $f$ corresponds to \elpiIn{con"f"}. 
This failure is motivated by the fact that \elpiIn{flam x\ fapp [fuva A, x]}
is a term belonging to \maybeeta.

@@

@@

\printAll
  {{\elpiIn{flam x\ fapp [fuva X, x] == flam x\ fapp [fuva X, x]}}}
  {{E0 == E2}}
  {{\mapping{X}{X'}{1}}}
  {{\linketaM{}{E2}{\lambda x.X'_{x}},
    \linketaM{}{E0}{\lambda x.X'_{x}}}}

@@

\printAll
  {{\elpiIn{flam x\ flam y\ fapp [fuva X, y, x] == flam x\ flam y\ x},
    \elpiIn{flam x\ fapp [fcon"f", fapp [fuva X, x], x] == fuva Y}}}
  {{E0 == \lambda x.\lambda y.x,
    E3 == E5}}
  {{\mapping{Y}{E5}{0},
    \mapping{X}{E2}{2}}}
  {{\linketaM{c0}{E4_{c0}}{\lambda x.E2_{c0 x}},
    \linketaM{}{E3}{\lambda x.(f~E4_{x}~x)},
    \linketaM{}{E0}{\lambda x.E1_{x}},
    \linketaM{c0}{E1_{c0}}{\lambda x.E2_{x c0}}}}
$$
{\arraycolsep=2pt \begin{array}{rrclrcl}
    \mathbb{P} = \{ &
       \lambda x.\lambda y.X \appsep y \appsep x & \Uo & \lambda x.\lambda y. x & \quad
       \lambda x. f \appsep (X \appsep x) \appsep x & \Uo&  Y ~\}\\
    \mathbb{T} = \{ &
       A &\Ue& \lambda x.\lambda y.x & \quad
       D &\Ue& F ~\}\\
    \mathbb{M} = \{ & \multicolumn{6}{l}{
      \mapping{Y}{F}{0} \quad
      \mapping{X}{C}{2} ~\}}\\
    \linkStore \;= \Big\{  &
      \multicolumn{6}{l}{ \begin{array}{ll}
          \linketaM{}{A}{\lambda x.B_{x}} & \quad
          \linketaM{x}{B_{x}}{\lambda y.C_{y x}}\\
          \linketaM{}{D}{\lambda x.(f~E_{x}~x)} & \quad
          \linketaM{y}{E_{y}}{\lambda w.C_{y w}}
      \end{array} ~\Big\}}
\end{array}}
$$
\marginpar{aggiustare i nomi delle variabili nel testo}
We see that the \elpiIn{maybe-eta} as detected $\lambda xy.F \appsep y \appsep
x$ and $\lambda x.f\appsep (F \appsep x) \appsep x$ and replaced them with
respectively the \Ho vars $X$ and $Z$. $X$ is linked with $\lambda x.Y_x$, $Y$
has arity $1$ and is $\eta$-linked with $\lambda y.H\appsep y \appsep x$ and $Z$
is linked to the term $\lambda x.f \appsep G_{x}~x$. However, the mapping
returned by the compilation, does not breaks \cref{inv:map-bijective}: the \Fo{}
variable $F$ is mapped to two different \Ho variables. To address this problem
and enforce \cref{inv:map-bijective}, we clean the mapping with a second phase
after the compilation. This phase is called \elpiIn{map-deduplication}.

@@

The complete problem to run for resolution is now:
%
\printAll
  {{}}
  {{X = \lambda x.\lambda y.x, 
    Z = f\appsep(\lambda x.x)}}
  {{\mapping{F}{H}{2}}}
  {{\linketaM{x}{Y_{x}}{\lambda y.H_{y x}},
    \linketaM{\phantom{x}}{X}{\lambda x.Y_{x}},
    \linketaM{\phantom{x}}{Z}{\lambda x.f \appsep G_{x}~x},
    \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}}}

>>>>

\include{pb_printer}

@@

\section{Handling of \maybeeta}\label{sec:eta}
\marginpar{TODO: fix syntax}
$\eta$-reduction is an equivalence relation where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expansion of $t$. 
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}
Following the implementation of \elpiIn{comp} given in
\cref{sec:compilation}, the unification problem $\mathbb{P}_1$ given 
above, is compiled into the unification problem $\mathbb{T}_1$. 
We can note that all the terms appearing in $\mathbb{P}_1$ are
unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathbb{T}_1$:
its left and right hand sides have both different rigid head:
$\lambda x.A'_x$ corresponds to the \Ho term \elpiIn{lam x\ app[con "f", x]} 
and $f$ corresponds to \elpiIn{con"f"}. 
This failure is motivated by the fact that \elpiIn{flam x\ fapp [fuva A, x]}
is a term belonging to \maybeeta.

@@

\printAlll
  {{{\lambda x.X \appsep x,f}}}
  {{{A,f}}}
  {{{X,B,1}}}
  {{{,A,\lambda x.B_x}}}

@@

\printAlll
  {{{\lambda x. X \appsep x, \lambda x. X \appsep x}}}
  {{{E0, E2}}}
  {{{X,A1,1}}}
  {{{,E2,\lambda x.A1_{x}},
    {,E0,\lambda x.A1_{x}}}}

@@

% \printAll
%   {{\elpiIn{flam x\ flam y\ fapp [fuva X, y, x] == flam x\ flam y\ x},
%     \elpiIn{flam x\ fapp [fcon"f", fapp [fuva X, x], x] == fuva Y}}}
%   {{E0 == \lambda x.\lambda y.x,
%     E3 == E5}}
%   {{\mapping{Y}{E5}{0},
%     \mapping{X}{E2}{2}}}
%   {{\linketaM{c0}{E4_{c0}}{\lambda x.E2_{c0 x}},
%     \linketaM{}{E3}{\lambda x.(f~E4_{x}~x)},
%     \linketaM{}{E0}{\lambda x.E1_{x}},
%     \linketaM{c0}{E1_{c0}}{\lambda x.E2_{x c0}}}}
$$
{\arraycolsep=2pt \begin{array}{rrclrcl}
    \mathbb{P} = \{ &
       \lambda x.\lambda y.X \appsep y \appsep x & \Uo & \lambda x.\lambda y. x & \quad
       \lambda x. f \appsep (X \appsep x) \appsep x & \Uo&  Y ~\}\\
    \mathbb{T} = \{ &
       A &\Ue& \lambda x.\lambda y.x & \quad
       D &\Ue& F ~\}\\
    \mathbb{M} = \{ & \multicolumn{6}{l}{
      \mapping{Y}{F}{0} \quad
      \mapping{X}{C}{2} ~\}}\\
    \linkStore \;= \Big\{  &
      \multicolumn{6}{l}{ \begin{array}{ll}
          \linketaM{}{A}{\lambda x.B_{x}} & \quad
          \linketaM{x}{B_{x}}{\lambda y.C_{y x}}\\
          \linketaM{}{D}{\lambda x.(f~E_{x}~x)} & \quad
          \linketaM{y}{E_{y}}{\lambda w.C_{y w}}
      \end{array} ~\Big\}}
\end{array}}
$$

\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{E0,\lambda x.\lambda y.x},
    {E3,E5}}}
  {{{Y,E5,0},
    {X,E4,1},
    {X,E2,2}}}
  {{{,E3,\lambda x.(f\appsep E4_{x}\appsep x)},
    {,E0,\lambda x.E1_{x}},
    {x,E1_{x},\lambda y.E2_{y x}}}}

\marginpar{aggiustare i nomi delle variabili nel testo}
We see that the \elpiIn{maybe-eta} as detected $\lambda xy.F \appsep y \appsep
x$ and $\lambda x.f\appsep (F \appsep x) \appsep x$ and replaced them with
respectively the \Ho vars $X$ and $Z$. $X$ is linked with $\lambda x.Y_x$, $Y$
has arity $1$ and is $\eta$-linked with $\lambda y.H\appsep y \appsep x$ and $Z$
is linked to the term $\lambda x.f \appsep G_{x}~x$. However, the mapping
returned by the compilation, does not breaks \cref{inv:map-bijective}: the \Fo{}
variable $F$ is mapped to two different \Ho variables. To address this problem
and enforce \cref{inv:map-bijective}, we clean the mapping with a second phase
after the compilation. This phase is called \elpiIn{map-deduplication}.

@@

The complete problem to run for resolution is now:
%
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{E0,\lambda x.\lambda y.x},
    {E3,E5}}}
  {{{Y,E5,0},
    {X,E2,2}}}
  {{{x,E4_{x},\lambda y.E2_{x y}},
    {,E3,\lambda x.(f\appsep E4_{x}\appsep x)},
    {,E0,\lambda x.E1_{x}},
    {x,E1_{x},\lambda y.E2_{y x}}}}
